---
title: "Migration on a network"
output:
  
  html_document:
    toc: true
    toc_float: true
    df_print: paged
  word_document: 
    toc: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(fig.width=7, fig.height=5)
options(max.print=23)
library(here);library(dplyr); library(stringr)
library(ggplot2)
library(tidyverse)
library(igraph)
```

# Introduction

Tigers have declined in numbers partly due to loss of habitat connectivity. Here, we use movement along a dynamic network to assess what habitat and connectivity characteristics facilitates or minimizes local extinctions. 

## Simple network

  - generate a simple network with 10 vertices fully connected.
  - assign population sizes of animals to each vertex
  - migration is possible through the network
  - migration is dependent on population size
  - add edge and vertex characteristics
    - Vertex can be vegetated, have food etc, have human habitaion
    - Edges can have weights associated with them due to roadways, forested networks, rivers/mountains, etc

```{r}
n_vertices <- 10

migration_network_historical <- make_full_graph(n = n_vertices, directed = FALSE)

plot(migration_network_historical, 
     vertex.color = "lightblue", 
     vertex.size = 30, 
     vertex.label.color = "black", 
     edge.color = "darkgray",
     main = "Fully Connected Migration Network")
```

### Vertex characteristics

We assign population sizes $p_i$ at each vertex i, with habitat characteristics such as habitat type $t_i$, human presence (hunting, conflict)  $h_i$, food availability  $f_i$,  climate  $c_i$, etc. $p_i$ is evolving with birth, death (due to carrying capacity $K$, life time), and migration. If $p_i = 0$ for a long time, make $K=0$.

```{r}


# Vertex characteristics 

#population sizes
V(migration_network_historical)$name <- paste0("population_",1:n_vertices)

V(migration_network_historical)$pop_size <- sample(50:500, 
                                                   size = vcount(migration_network_historical), replace = TRUE)


# habitat type
habitat_types <- c("Old Growth", "Wetland", "Shrubland", "Secondary Forest")


V(migration_network_historical)$habitat <- sample(habitat_types, 
                                                size = vcount(migration_network_historical), replace = TRUE)

#human presence

human_presence <- c("High", "Medium", "Low", "None")

V(migration_network_historical)$human_presence <- sample(human_presence, 
       size = vcount(migration_network_historical), replace = TRUE, 
       prob = c(0.2, 0.3, 0.3, 0.2))

# food availability 

food_availability <- c("High", "Medium", "Low", "None")

V(migration_network_historical)$food_availability <- sample(food_availability, 
       size = vcount(migration_network_historical), replace = TRUE, 
       prob = c(0.3, 0.4, 0.2, 0.1))

#plot
plot(migration_network_historical, 
     vertex.label.cex = 0.8,
## Add Edge characteristics= "black",
     vertex.size = V(migration_network_historical)$pop_size/20
     )

```



### Edge characteristics

We assign corridor types to edges, which can be wildlife corridor (for example - dense forest, open woodland, etc), terrain (river crossing, mountains, desert plateau), or urbanization (highways, human settlements). Each corridor type has a weight associated with it that affects migration.

Friction maps available for terrain. That can influence edge weight.


```{r}
corridor_types <- c("Dense Forest", "Open Woodland", "River Crossing", "Highway Obstruction")

E(migration_network_historical)$corridor_type <- sample(corridor_types, 
                                        size = ecount(migration_network_historical), 
                                        replace = TRUE,
                                        prob = c(0.3, 0.4, 0.15, 0.15))


weight_lookup <- c(
  "Dense Forest"        = 0.9,
  "Open Woodland"       = 0.7,
  "River Crossing"      = 0.3,
  "Highway Obstruction" = 0.1
)

E(migration_network_historical)$weight <- weight_lookup[E(migration_network_historical)$corridor_type]

#visualize the network
# Map habitat types to distinct colors
habitat_colors <- c(
  "Old Growth"       = "#2e7d32",
  "Secondary Forest" = "#81c784",
  "Wetland"          = "#0288d1",
  "Shrubland"        = "#fbc02d"
)
V(migration_network_historical)$color <- habitat_colors[V(migration_network_historical)$habitat]

#plot
plot(migration_network_historical, 
     vertex.label.cex = 0.8,
     vertex.label.color = "black",
     vertex.size = V(migration_network_historical)$pop_size/20,
     edge.width = E(migration_network_historical)$weight*5
     )

legend(
  "bottomleft",
  legend = names(habitat_colors),
  fill = habitat_colors,
  title = "Habitat",
  bty = "n",
  cex = 0.8
)



```


### Assign carrying capacity

Based on habitat type, food availability, and human presence. 

```{r}
# assign carrying capacity

base_capacity_lookup <- c(
  "Old Growth"       = 600,
  "Wetland"          = 450,
  "Secondary Forest" = 300,
  "Shrubland"        = 150
)

# Capacity retention modifier under human pressure
human_impact_lookup <- c(
  "None"    = 1.00,  # Negligible impact
  "Low"     = 0.85,  # 15% reduction in capacity
  "Medium" = 0.70,  # 30% reduction in capacity
  "High"   = 0.35   # 65% reduction due to high disturbance
)

food_availability_lookup <- c(
  "High"    = 1.00,  # Full capacity
  "Medium"  = 0.90,  # 10% reduction in capacity
  "Low"     = 0.75,  # 25% reduction in capacity
  "None"    = 0.50   # 50% reduction due to lack of food
)

# computing capcity
V(migration_network_historical)$capacity <- round(base_capacity_lookup[V(migration_network_historical)$habitat] * human_impact_lookup[V(migration_network_historical)$human_presence] * 
food_availability_lookup[V(migration_network_historical)$food_availability])

V(migration_network_historical)$pop_size <- round(V(migration_network_historical)$capacity*runif(vcount(migration_network_historical), min = 0.8, max = 1.1))

igraph::as_data_frame(migration_network_historical, what = "vertices")[, c("name", "habitat", "food_availability", 
                                       "human_presence", "capacity", "pop_size")]



```



### Migration, Birth, Death

```{r}



simulate_migration_step <- function(graph, m0 = 0.2){
  
  n <- vcount(migration_network_historical)
  current_pop <- V(migration_network_historical)$pop_size
  capacities <- V(migration_network_historical)$capacity
  
  
  adj_weights <- as.matrix(as_adjacency_matrix(migration_network_historical, attr = "weight", sparse = FALSE))
  
  density <- current_pop / capacities
  
  p_out <- pmin(m0*density, 0.6)
  
  #migration transition matrix
  
  T_mat <- matrix(0, nrow = n, ncol = n)
  
  for(i in seq_len(n)){
    total_weight <- sum(adj_weights[i,])
    
    if(total_weight > 0){
      T_mat[i,] <- (adj_weights[i,]/total_weight)*p_out[i]
    }
  }
  
  
  #migration matrix
  
  migration_mat <- matrix(0, nrow = n, ncol = n)
  
  for(i in seq_len(n)){
    for(j in seq_len(n)){
      if(i != j){
        migration_mat[i,j] <- round(current_pop[i]*T_mat[i,j])
      }
    }
  }
  
  
  emigrants <- rowSums(migration_mat)
  immigrants <- colSums(migration_mat)
  
  V(migration_network_historical)$pop_size <- current_pop - emigrants + immigrants
  return(list(graph = migration_network_historical, flux = migration_mat ))
  
}


# simulate over generations

generations <- 20
pop_history <- matrix(NA, nrow = generations, ncol = vcount(migration_network_historical))

colnames(pop_history) <- V(migration_network_historical)$name


for(gen in 1:generations){
  pop_history[gen, ] <- V(migration_network_historical)$pop_size
  step_result <- simulate_migration_step(migration_network_historical, m0 = 0.2)
  migration_network_historical <- step_result$graph
}

# plot population sizes as stacked plot
pop_df <- as.data.frame(pop_history)
pop_df$generation <- seq_len(nrow(pop_df))

pop_long <- pivot_longer(pop_df, cols = -generation, names_to = "Population", values_to = "Size")

pop_long$Population <- factor(pop_long$Population, levels = V(migration_network_historical)$name)

ggplot(pop_long, aes(x = generation, y = Size, fill = Population)) +
  geom_area(alpha = 0.85, color = "white", linewidth = 0.25) +
  scale_fill_viridis_d(option = "turbo") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "",
    x = "Simulation Generation",
    y = "Total Number of Individuals",
    fill = "Subpopulation"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "none" # Combined into the network legend
  )

plot(
  migration_network_historical,
  # layout = net_layout,
  vertex.size = V(migration_network_historical)$pop_size / 25,       # Node size scaled to final population size
  
  vertex.label = V(migration_network_historical)$name,
  vertex.label.cex = 0.7,
  vertex.label.color = "black",
  edge.width = E(migration_network_historical)$weight * 4,           # Line width scaled to corridor weight
  edge.color = "gray60",
  main = "Final Network State"
)

```









